Large-scale neural circuits on an FPGA

### 2015-05-14 docs/paper/paper.tex
fixed multiplier figures

\documentclass[english]{article}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{babel}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{listings}

@@

@@

\subsection{Potential Applications}

@@

@@

\subsection{Description of NEF and population mode}

@@

@@

\begin{subfigure}[b]{0.5\textwidth}
\centering
\includegraphics[width=3in]{multiplier-sw.eps}
\caption{Software}
\label{fig:multiplier:sw}
\end{subfigure}

\begin{subfigure}[b]{0.5\textwidth}
\centering
\includegraphics[width=3in]{multiplier.eps}
\caption{Hardware}
\label{fig:multiplier:hw}
\end{subfigure}

@@

>>>>

\documentclass[english]{article}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{babel}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{listings}
\usepackage{authblk}
\usepackage{amsmath}

@@

\title{Large-scale neural circuits on an FPGA}

\author[1]{Murphy Berzish}
\author[1,2,3]{Chris Eliasmith}
\author[1,2]{Bryan Tripp}
\affil[1]{Centre for Theoretical Neuroscience, University of Waterloo}
\affil[2]{Dept. of Systems Design Engineering, University of Waterloo}
\affil[3]{Dept. of Philosophy, University of Waterloo}
\maketitle

\abstract{abstract goes here - bryan???}

@@

- potential applications

@@

- from my other paper for mind collection

@@

\subsection{The Neural Engineering Framework}

The Neural Engineering Framework (NEF) is method for constructing biologically realistic neural models \cite{Eliasmith1999s}.  Several overviews of the NEF are available \cite{Eliasmith2003f,Eliasmith2005p,stewart2012b,Eliasmith2012b}, but the description provided here is based on that in \cite{Stewart2014}.   
The NEF is a general-purpose approach to
implementing algorithms using components such
as spiking neurons \cite{Eliasmith2003m}. As a result, the NEF can be thought of
as a ``neural compiler'' where algorithms characterized by a high-level
mathematical description are converted into neural networks.
This compilation process works for arbitrary neuron types, and can
be constrained in biologically informed ways. Importantly, the high-level
description must be expressed in terms of vectors and functions on
those vectors (including differential equations). The resulting
neural networks approximate the desired functions, and the error of
this approximation can be made small by increasing the
number of neurons.

While the NEF can be used to build arbitrary abstract systems such
as controlled attractor networks \cite{Eliasmith2005p}, we have primarily
used it to show how particular capabilities found in real animals
might be implemented biologically. This has included path integration
in rodents \cite{Conklin2005b}, working memory \cite{Singh2006b}
and arm movements \cite{Dewolf} in monkeys, and decision-making in
rats \cite{Laubach2010,Liu2011} and humans \cite{Litt2008u}. We
have also taken into account biological constraints such as Dale's
Principle \cite{Parisien2008c} and incorporated biologically realistic
learning rules to construct these networks \cite{MacNeil2011a,bekolay2013}.

The remainder of this section serves as a summary of the three main principles
of the NEF.

\subsection{Principle 1 - Representation}

In the NEF, groups of neurons are taken to represent vectors,
and connections between groups of neurons compute functions on those
vectors. The first NEF principle shows how the activity of a group
of neurons can be said to represent a vector, and how changes in the
activity of those neurons corresponds to changes in the represented
vector.

In the brain many neurons have a particular stimulus (or response) for
which they will fire most strongly. As an input stimulus (or response)
changes to become less similar to the preferred stimulus, the neuron
will fire less quickly.  This kind of behaviour was originally identified in the motor
system \cite{Georgopoulos1989q} and has since been seen in the head
direction system \cite{Taube2007}, visual system \cite{Rust2006},
and auditory system \cite{Fischer2009w}. The NEF identifies the preferred stimulus with a ``preferred direction vector'' associated with each neuron.  For a more detailed exploration of this idea, see \cite{Stewart2011a}.

In the NEF, we generalize this concept to all neural populations (those that are stimulus driven and those that are not).
In particular, we quantify it by stating that the total current going
into a neuron will be proportional to the dot product of the vector
to be represented, $\mathbf{x}$, and the preferred direction (or ``encoding'') vector
for the neuron, $\mathbf{e}_{i}$ (plus a constant bias term). The
response of a neuron \textbf{$i$} for any given input vector \textbf{$\mathbf{x}$}
is thus

\begin{equation}
\delta_{i}(\mathbf{x})=G_{i}[\alpha_{i}\mathbf{e}_{i}\mathbf{x}+J_{i}^{bias}]\label{eq:enc}
\end{equation}
where $\delta_{i}$ is the spiking output of the neuron, $G_{i}$
is the neuron model, $\alpha_{i}$ is a randomly chosen gain term,
$\mathbf{e}_{i}$ is the preferred direction vector, and $J_{i}^{bias}$
is a randomly chosen fixed background current. We use $\mathbf{e}$
for \textit{encoder} to indicate that (\ref{eq:enc}) captures a transformation
between spaces: \textbf{$\mathbf{x}$} is encoded in the activity
space of the neurons.  Importantly, the NEF applies
to a large variety of neuron models (including both spiking and non-spiking
models) since it makes no commitment to a specific function $G$ (whose
input is the total current flowing into the neuron). The leaky-integrate-and-fire
(LIF) model is a common choice, for reasons of computational efficiency, but a wide variety of
neural models work with the NEF.

Given an encoding operation, it is natural to define a \textit{decoding} operation,
in order to characterize the information processing characteristics
of the system (in this case an ensemble of neurons). To decode a continuous
estimate of the input from neural activity, the NEF focuses on the
postsynpatic filtered activity generated by the reception of a spike
at a synapse. This activity is taken to be a linear filter applied
to the spiking activity:
\[
a_{i}(\mathbf{x})=\sum_{j}h_{i}(t)*\delta_{i}(t-t_{j}(\mathbf{x}))
\]
where $h_{i}(t)$ is the synaptic response function (usually a decaying
exponential) with a time constant, $\tau_{PSC}$, determined by neurotransmitter
type at the synapse, '$*$' is the convolution operator, and $\delta_{i}(t-t_{j}(\mathbf{x}))$
is the spike train produced by neuron $i$ in response to input $\mathbf{x}$,
with spike times indexed by $j$. Having defined this continuous variable,
which is equal to the spike rate as $\tau_{PSC}\rightarrow\infty$,
we can specify a decoding operation for estimating the input $\mathbf{x}$.

For reasons that will be apparent in a moment, we use a \textit{linear}
decoder:

\begin{equation}
\hat{\mathbf{x}}=\sum_{i}^{N}a_{i}\mathbf{(x)}\mathbf{d}_{i}\label{eq:dec}
\end{equation}
where $N$ is the number of neurons in the group, $\mathbf{d}_{i}$
are the linear decoders, and $\hat{\mathbf{x}}$ is the estimate of
the original $\mathbf{x}$ value that produced the neural activity
(\ref{eq:enc}).

Any optimization method can be used to find these decoders. The simplest
is to use standard least-squares optimization:
\begin{equation}
\underset{\mathbf{d}_{i}}{arg\, min}\int[\mathbf{x}-\sum_{i}^{N}a_{i}(\mathbf{x)}\mathbf{d}_{i}]^{2}d\mathbf{x}\label{eq:error}
\end{equation}
where $\mathbf{d}_{i}$ are the decoding vectors over which this error
is minimized and the integral is over all $\mathbf{x}$ values. It
has been shown that linear decoders are sufficient to decode \textasciitilde{}95\%
of the information available in a single spike train generated by
a stimulus \cite{Rieke1997b}. Furthermore, as the number of neurons
$N$ increases, the mean squared error decreases as $1/N$. The NEF
decoding process is depicted in Figure \ref{fig:NEF-decoding}, where
the optimal linear decoders have been found and used for twenty neurons.
Temporal decoding is also linear, as described, and is performed using
the postsynaptic, current-based filters (see section \ref{sub:Principle-3--}
for further discussion).

While linear decoders are useful for visualizing the information encoded
within the activity of a group of neurons, they also provide a direct
way to solve for connection weights between groups of neurons. This
is a key advantage of the NEF: rather than using a learning rule to
optimize over the entire space of all connection weights, we instead
solve the simpler problem of optimizing over the space of linear decoders,
and then use that result to solve for the connection weights. Importantly,
given the characterization of neural activity with preferred direction
vectors, there is no difference between using this smaller space and
using the equivalent full connection matrix. Interestingly, this general
technique has started to be applied in broader domains; see \cite{tapson2013}
for an overview.

For example, if a connection between neural groups is meant to compute
the identity function $\mathbf{y}=\mathbf{x}$ (where $\mathbf{y}$
is the vector space represented by the second population B and $\mathbf{x}$
is the vector space represented by the first population A), the connections
between individual neurons are given by 
\begin{equation}
\omega_{ij}=\alpha_{j}\mathbf{d}_{i}\mathbf{e}_{j}\label{eq:weights}
\end{equation}
where $i$ indexes the neurons in group A and $j$ indexes the neurons
in B. Behavior of this network is shown in Figure \ref{fig:Using-NEF-weights}a.

While the least-squares method for optimization we use here is not
biologically plausible on its own, we have also shown that biologically
realistic learning rules will converge on a similar solution \cite{MacNeil2011a}.
These realistic rules are, however, several orders of magnitude more
computationally expensive.

\subsection{Principle 2 - Transformation}

Connections between groups of neurons can also compute functions other
than the identity function. That is instead of \textbf{$\mathbf{y}=\mathbf{x}$}
we can do \textbf{$\mathbf{y}=f(\mathbf{x})$}. We do this by finding
decoders \textbf{$\mathbf{d}_{i}^{f}$ }for the particular function\textbf{
$f(\mathbf{x})$} by substituting \textbf{$f(\mathbf{x})$ }for $\mathbf{x}$
in equation (\ref{eq:error}).

\begin{equation}
\underset{\mathbf{d}_{i}^{f}}{arg\, min}\int[f(\mathbf{x})-\sum_{i}a_{i}(\mathbf{x)}\mathbf{d}_{i}^{f}]^{2}d\mathbf{x}.\label{eq:error-fcns}
\end{equation}

The connection weights can then be computed using (\ref{eq:weights}).
For the special case of linear functions\textbf{ $\mathbf{y}=\mathbf{L}\mathbf{x}$},
rather than solving for a new decoder, we can simply put $\mathbf{L}$
directly into the weight equation itself, resulting in $\omega_{ij}=\alpha_{j}\mathbf{d}_{i}\mathbf{L}\mathbf{e}_{j}$.
Combining these two approaches, the neural connection weights needed
to approximate the function\textbf{ $\mathbf{y=L}f(\mathbf{x\textrm{)}}$}
are:
\begin{equation}
\omega_{ij}=\alpha_{j}\mathbf{d}_{i}^{f}\mathbf{L}\mathbf{e}_{j}\label{eq:weights-general}
\end{equation}
Figure \ref{fig:Using-NEF-weights}b shows the computation of the
element-wise square ($f(\mathbf{x})=[x_{1}^{2},x_{2}^{2}]$).

\subsection{\label{sub:Principle-3--}Principle 3 - Dynamics}

While the first two principles are sufficient to build neural approximations
of any desired function of the vector $\mathbf{x}$, the NEF also
provides a method for computing functions of the form \textbf{$\mathbf{\frac{\mathrm{d}x}{\mathrm{dt}}=}f(\mathbf{x},\mathbf{u})$},
where $\mathbf{u}$ is the input from some other population. We do
this by exploiting the fact that neurons do not simply accept input
as spikes. Rather, when a spike is transmitted from one neuron to
another, the actual current that flows into the second neuron is a
low-pass-filtered version of that spike. In particular, the post-synaptic
current is well-approximated by $h(t)=u(t)e^{-t/\tau}$, where $u(t)$
is the step function and $\tau$ is the time constant of the neurotransmitter
used. This time constant varies throughout the brain, e.g., from 2-5
ms (AMPA; \cite{jonas1993quantal}) up to \textasciitilde{}100ms (NMDA;
\cite{sah1990properties}). The effect of this filter is that instead
of a connection computing the function $\mathbf{y}(t)=f(\mathbf{x}(t))$
it will compute $\mathbf{y}(t)=f(\mathbf{x}(t))*h(t)$ (or, in the
Laplace domain, $\mathbf{Y}(s)=\mathbf{F}(s)H(s)$).

It turns out that this implicit low-pass filter can be used to generate
neural models that compute abitrary dynamics. Given a neural population
representing $\mathbf{x}$, an input $\mathbf{u}(t)$, and a feedback
connection from $\mathbf{x}$ back to itself computing $g(\mathbf{x}(t))$,
we note that in the Laplace domain we get $\mathbf{X}(s)=(\mathbf{G}(s)+\mathbf{U}(s))H(s)$.
Since the Laplace transform of $h(t)=u(t)e^{-t/\tau}$ is $H(s)=1/(1+s\tau)$,
we can rearrange this to get $s\mathbf{X}(s)=(\mathbf{G}(s)-\mathbf{X}(s))/\tau+\mathbf{U}(s)/\tau$.
Converting back to the time domain, \textbf{$\mathbf{\frac{\mathrm{d}x}{\mathrm{dt}}=}\frac{g(\mathbf{x}(t))-\mathbf{x}(t)}{\tau}+\frac{\mathbf{u}(t)}{\tau}$}.
Thus, if we desire the dynamics \textbf{$\mathbf{\frac{\mathrm{d}x}{\mathrm{dt}}=}f(\mathbf{x}(t))+\mathbf{u}(t)$},
we introduce a feedback connection that uses the previous two NEF
principles to find connection weights that compute $g(\mathbf{x}(t))=\tau f(\mathbf{x})+\mathbf{x}$
and we scale the input $\mathbf{u}(t)$ by $\tau$. For arbitrary
dynamics of the form \textbf{$\mathbf{\frac{\mathrm{d}x}{\mathrm{dt}}=}f(\mathbf{x},\mathbf{u})$},
we do a change of variables $\mathbf{x}^{\prime}=\langle\mathbf{x},\mathbf{u}\rangle$.

This exploitation of the inherent first-order low-pass filter found
in synaptic connections allows for the implementation of a very wide
variety of systems, including oscillators, integrators, and arbitrary
attractor networks \cite{Eliasmith2005p}. Figure \ref{fig:Two-dynamical-systems}a
shows a single neural population with a recurrent feedback connection
computing the standard linear oscillator \textbf{$\mathbf{\frac{\mathrm{d}x}{\mathrm{dt}}=}[-x_{2},-x_{1}]$}.
Here, the post-synapic time constant $\tau$ is 100ms, but the period
of oscillation is $2\pi$ (\textasciitilde{}6.28 seconds). Importantly,
we achieve different periods of oscillation without changing the neural
property $\tau$; rather, we change the function being computed by
the feedback connection, resulting in a different set of connection
weights. For example, we could compute \textbf{$\mathbf{\frac{\mathrm{d}x}{\mathrm{dt}}=}[-x_{2}/2,-x_{1}/2]$},
giving a period of $\pi$, instead. This same method works for nonlinear
functions as well; in Figure \ref{fig:Two-dynamical-systems}b we
show the classic Lorenz attractor \textbf{$\mathbf{\frac{\mathrm{d}x}{\mathrm{dt}}=}[10(x_{2}-x_{1}),x_{1}(28-x_{3})-x_{2},x_{1}x_{2}-\frac{8}{3}x_{3}]$}.
This approach allows for the construction of neural models that correspond
to a very large family of functions, including those typically employed
by modern control theory and dynamic systems theory.

It should be noted that networks constructed with this approach are
fundamentally different from Echo State Networks \cite{jaeger2001}
and Liquid State Machines \cite{Maass2002}. In both of those approaches,
the recurrent connections are randomly chosen, and then a linear combination
of the outputs are found that compute the desired function. With the
NEF, we also find a linear combination of the outputs ($\mathbf{d}_{i}$),
but we solve for the ideal recurrent connection weights to achieve
the desired dynamics. This makes the NEF much more efficient and capable
of computing a much wider range of dynamics.

\subsection{Nengo}

The principles of the NEF have been incorporated into the Nengo neural simulator \cite{Bekolay2014}.  This simulator

- mention spaun

These three principles are sufficient to implement all of our neural
models. However, to simplify the process of constructing these models,
we have developed an open-source software package known as Nengo (Neural
ENGineering Objects) that creates and runs these models. Models can
be built in Nengo using a drag-and-drop graphical user interface or
specified using the Python scripting language. Full details and documentation
can be found online at http://nengo.ca, and in other publications
\cite{Stewart2009l}.

For example, to create the model shown in Figure \ref{fig:Using-NEF-weights}a,
we use the following script:
\begin{verbatim}
net = nef.Network("Identity Function")
net.make("A", neurons=20, dimensions=1) 
net.make("B", neurons=20, dimensions=1)
net.connect("A", "B") 
\end{verbatim}

For Figure \ref{fig:Using-NEF-weights}b we need to compute the element-wise
square. This is specified in a Python function as follows:

\begin{verbatim}
net = nef.Network("Element-wise Square")
net.make("A", neurons=20, dimensions=1) 
net.make("B", neurons=20, dimensions=1)

def square(x):
    return x{[}0{]}{*}x{[}0{]}, x{[}1{]}{*}x{[}1{]}

net.connect("A", "B", func=square) 
\end{verbatim}

Nengo will automatically solve for the connection weights that will
best approximate the provided function.

Nengo also provides an interactive interface for displaying the results
of a simulation while it is running, allowing for real-time interaction
with a running model. This interface allows the generated plots to
be exported, and was used to produce the previous figures in this
paper (i.e. all figures in section \ref{sec:The-Neural-Engineering}).
Furthermore, Nengo scales up to our largest models: the 2.5 million
neuron Spaun model is run in Nengo, and can be downloaded at http://models.nengo.ca/spaun.

- backends

\subsubsection{Population mode}

- Bryan???

@@

- Murphy???

@@

\begin{subfigure}[b]{0.4\textwidth}
\centering
\includegraphics[width=4in]{multiplier-sw.eps}
\caption{Software}
\label{fig:multiplier:sw}
\end{subfigure}

\begin{subfigure}[b]{0.4\textwidth}
\centering
\includegraphics[width=4in]{multiplier.eps}
\caption{Hardware}
\label{fig:multiplier:hw}
\end{subfigure}

@@

\bibliographystyle{acm}
\bibliography{references}
